---
title: "Assignment 2: Coding Basics"
author: "Natalie von Turkovich"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on coding basics. 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your first and last name into the file name (e.g., "FirstLast_A02_CodingBasics.Rmd") prior to submission.



## Basics Day 1

1. Generate a sequence of numbers from one to 100, increasing by fours. Assign this sequence a name. 

2. Compute the mean and median of this sequence. 

3. Ask R to determine whether the mean is greater than the median. 

4. Insert comments in your code to describe what you are doing.

```{r}
#1. Finding a sequence from 1 to 100, at intervals of 4. Assigning this function to the name sequence1.
sequence1<-seq(1,100,4)
sequence1

#2. Using the mean and median function to find the mean and median of sequence1. 
mean(sequence1)
median(sequence1)

#3. Using a conditional statement to find if the mean is bigger than the median of sequence1. The statement came back false, the mean is not bigger. 
mean(sequence1)>median(sequence1)

```

## Basics Day 2

5. Create a series of vectors, each with four components, consisting of (a) names of students, (b) test scores out of a total 100 points, and (c) whether or not they have passed the test (TRUE or FALSE) with a passing grade of 50.

6. Label each vector with a comment on what type of vector it is.

7. Combine each of the vectors into a data frame. Assign the data frame an informative name.

8. Label the columns of your data frame with informative titles.

```{r}
#5. 
names<- c("John", "Paul", "George", "Ringo")
names #character vector

test_scores<- c(85, 45, 95, 100)
test_scores #integer vector

pass_fail<- test_scores>50
pass_fail #logical vector

#7
class_grades.df<- data.frame(names, test_scores, pass_fail)
class_grades.df

#8
colnames(class_grades.df) <- c('Student Names','Score','Passed Class')
class_grades.df

```

9. QUESTION: How is this data frame different from a matrix?

> Answer: The difference between a data frame and a matrix is that a martix can only contain one class of data and a data frame can contain many different classes of data. 

10. Create a function with an if/else statement. Your function should determine whether a test score is a passing grade of 50 or above (TRUE or FALSE). You will need to choose either the `if` and `else` statements or the `ifelse` statement. Hint: Use `print`, not `return`. The name of your function should be informative.

11. Apply your function to the vector with test scores that you created in number 5. 

```{r}
#10
passing_grade<-function(x){ifelse(x>50, "pass", "fail")}

#11
passing_grade(test_scores)

```

12. QUESTION: Which option of `if` and `else` vs. `ifelse` worked? Why?

> Answer: The ifelse function worked well in this case. It is a more concise way to code using a logical expression to decide the output of the function. I believe you could get the same result with the if and else functions used separately but it would not be as effcient, it would use more lines of code. 


