---
title: "Assignment 5: Data Visualization"
author: "Natalie von Turkovich"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1
getwd()
library(tidyverse)
library(cowplot)
library(lubridate)

PeterPaul.chem.nutrients <-
  read.csv("/Users/natalievonturkovich/Documents/DUKE/Courses/Spring 22/ENV_872_EDA/Environmental_Data_Analytics_2022/Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed_NVT.csv", 
           stringsAsFactors = TRUE)
PeterPaul.chem.nutrients.gathered <-
  read.csv("/Users/natalievonturkovich/Documents/DUKE/Courses/Spring 22/ENV_872_EDA/Environmental_Data_Analytics_2022/Data/Processed/NTL-LTER_Lake_Nutrients_PeterPaulGathered_Processed_NVT.csv", 
           stringsAsFactors = TRUE)
Niwot.litter.processed <-
  read.csv("/Users/natalievonturkovich/Documents/DUKE/Courses/Spring 22/ENV_872_EDA/Environmental_Data_Analytics_2022/Data/Processed/NEON_NIWO_Litter_mass_trap_Processed_NVT.csv", 
           stringsAsFactors = TRUE)

#2 
PeterPaul.chem.nutrients$sampledate <- ymd(PeterPaul.chem.nutrients$sampledate)
PeterPaul.chem.nutrients.gathered$sampledate <- ymd(PeterPaul.chem.nutrients.gathered$sampledate)
Niwot.litter.processed$collectDate <- ymd(Niwot.litter.processed$collectDate)

```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.position = "right") #alternative: legend.position + legend.justification

theme_set(mytheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
NTL_LITER_4 <-
  ggplot(PeterPaul.chem.nutrients, aes(x = tp_ug, y = po4, color = lakename))+
  geom_point() +
  geom_smooth(method = lm, color = "black") +
  xlim(0, 70) +
  ylim(0, 15)+
  labs( x = "Total Phosphorus", y = "Phosphate", color = "Lake Name", 
        title = "Peter & Paul Lakes - Phosphorus and Phosphate")
print(NTL_LITER_4)
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5
PeterPaul.chem.nutrients$month<-as.factor(PeterPaul.chem.nutrients$month)

NTL_LITER_5_temp <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = temperature_C)) +
  geom_boxplot(aes(color = lakename)) +
  labs(y = "Temperature (C)", x = "Month", color = "Lake Name") 
print(NTL_LITER_5_temp)

NTL_LITER_5_tp <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tp_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(y = "TP", x = "Month", color = "Lake Name")
print(NTL_LITER_5_tp)

NTL_LITER_5_tn <-
  ggplot(PeterPaul.chem.nutrients, aes(x = month, y = tn_ug)) +
  geom_boxplot(aes(color = lakename)) +
  labs(y = "TN", x = "Month", color = "Lake Name")
print(NTL_LITER_5_tn)

plot_5<-plot_grid(NTL_LITER_5_temp + theme(legend.position="none"), 
                  NTL_LITER_5_tp + theme(legend.position="none"), 
                  NTL_LITER_5_tn + theme(legend.position="none"), 
                  nrow = 1, align = 'h', rel_heights = c(1, 1,1))

legend_5 <- get_legend(
  NTL_LITER_5_temp + 
    guides(color = guide_legend(nrow = 1)) +
    theme(legend.position = "bottom"))

title_5 <- ggdraw() + draw_label("Peter & Paul Lake Temperature, TP & TN", fontface='bold') 

plot_grid(title_5, plot_5, legend_5, ncol = 1, rel_heights = c(.1, 1, .1))
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Over seasons, temperature has the widest variabillity, where as TP and even more so TN have a smaller range of where most observations sit, with a larger number of outliers. Between lakes, Peter lake has a wider range of temperature readings for each month. For TP and TN, Peter lake also has a wider range, but in general Peter Lake values for these characteristics are higher than Paul Lake. 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

```{r}
#6
Needles_plot <-
  ggplot(filter(Niwot.litter.processed, functionalGroup == "Needles") , 
         aes(x = collectDate, y = dryMass)) +
  geom_point(aes(color = nlcdClass)) +
  labs(y = "Dry Mass", x = "Collection Date", color = "NLCD Class", 
       title = "Needle Dry Mass by Month") +
  #scale_x_date(date_breaks = "2 months", date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))
print(Needles_plot)
```

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#7
Needles_plot_facet <-
  ggplot(filter(Niwot.litter.processed, functionalGroup == "Needles") , 
         aes(x = collectDate, y = dryMass)) +
  geom_point(aes(color = nlcdClass)) +
  labs(y = "Dry Mass", x = "Collection Date", color = "NLCD Class", 
       title = "Needle Dry Mass by Month") +
  #scale_x_date(date_breaks = "2 months", date_labels = "%b %y") +
  theme(axis.text.x = element_text(angle = 45,  hjust = 1))+
  facet_wrap(vars(nlcdClass), nrow = 3)
print(Needles_plot_facet)
```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I ghink that plot 7 is more effective than plot 6. Plot 7 separates out each NLCD class, making it easier to see the distribuation of dry mass for each class. You are able to see what the distribution of each class looks like then compare them to eachother. With all of the classes in the same plot it is hard to see how they compare with all of the points overlapped. 
